Orienting the causal relationship between imprecisely measured traits using GWAS summary data

### 2017-07-18 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Orienting the causal relationship between imprecisely measured traits using genetic instruments},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using genetic instruments}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{16 July 2017}

@@

Inference of the causal structure that induces correlations between two
traits can be achieved by combining genetic associations with a
mediation-based approach, as is done in the causal inference test (CIT)
and others. However, we show that measurement error in the phenotypes
can lead to mediation-based approaches inferring the wrong causal
direction, and that increasing sample sizes has the adverse effect of
increasing confidence in the wrong answer. Here we introduce an
extension to Mendelian randomisation, a method that uses genetic
associations in an instrumentation framework, that enables inference of
the causal direction between traits, with some advantages. First, it is
less susceptible to bias in the presence of measurement error; second,
it is more statistically efficient; third, it can be performed using
only summary level data from genome-wide association studies; and
fourth, its sensitivity to measurement error can be evaluated. We apply
the method to infer the causal direction between DNA methylation and
gene expression levels. Our results demonstrate that, in general, DNA
methylation is more likely to be the causal factor, but this result is
highly susceptible to bias induced by systematic differences in
measurement error between the platforms. We emphasise that, where
possible, implementing MR and appropriate sensitivity analyses alongside
other approaches such as CIT is important to triangulate reliable
conclusions about causality.

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding (1--3). In response to this, the use of genetic
associations to instrument traits has emerged as a technique for
improving the reliability of causal inference in observational data, and
with the coincident rise in genome-wide association studies it is now a
prominent tool that is applied in several different guises (3--6).
However, potential pitfalls remain and one that is often neglected is
the influence of non-differential measurement error on the reliability
of causal inference.

@@

Measurement error will of course arise in other types of data too. For
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect (12). A similar problem of
biological misspecification is unavoidable in disease diagnosis, and
measuring behaviour such as smoking or diet is notoriously difficult to
do accurately. Measurement error can also be introduced after the data
have been collected, for example the transformation of non-normal data
for the purpose of statistical analysis will lead to a new variable that
will typically incur both changes in scale and imprecision (noise)
compared to the original variable. The sources of measurement error are
not limited to this list (8), and its impact has been explored in the
epidemiological literature extensively (13,14). Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

@@

MR can be applied to the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP is used as a surrogate for the exposure. Assuming the SNP
associates with the outcome only through the exposure, the causal effect
of the exposure on the outcome can be estimated by scaling the
association between the SNP and the outcome by the association between
the SNP and the exposure. Though difficult to test empirically, this
assumption can be relaxed in various ways when multiple instruments are
available for a putative exposure (24,25) and a number of sensitivity
tests are now available to improve reliability (26).

@@

MR has some important advantages over the mediation-based approaches.
First, the mediation-based approaches require that the exposure, outcome
and instrumental variables are all measured in the same data, whereas
recent extensions to MR circumvent this requirement, allowing causal
inference to be drawn when exposure variables and outcome variables are
measured in different samples (28). This has the crucial advantage of
improving statistical power by allowing analysis in much larger sample
sizes, and dramatically expands the breadth of possible phenotypic
relationships that can be evaluated (26). Second, the mediation-based
approach of adjusting the outcome for the exposure to nullify the
association between the SNP and the outcome is affected by unmeasured
confounding of the exposure and outcome. This is because adjusting the
outcome by the exposure induces a collider effect between the SNP and
outcome (29), and the in order to fully abrogate this association one
must also adjust for all (hidden or otherwise) confounders. MR does not
suffer from this problem because it does not test for association
through adjustment. Third, when MR assumptions are satisfied the method
is robust to there being measurement error in the exposure variable
(30). Indeed instrumental variable (IV) analysis was in part initially
introduced as a correction for measurement error in the exposure (31),
whereas it has been noted that both classic mediation-based analyses
(13,14,32,33) and mediation-based methods that use instrumental
variables (34,35) are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. We then present an extension to MR that
allows researchers to ascertain the causal direction of an association
even when the biology of the instruments are not fully understood, and
also a metric to evaluate the sensitivity of the result of this
extension to measurement error. Together these extensions improve the
utility of MR in cases where mediation based methods might have
otherwise been used preferentially. Finally, we apply this method to
infer the direction of causation between DNA methylation levels and gene
expression levels. Our analyses highlight that because these different
causal inference techniques have varying strengths and weaknesses,
triangulation of evidence from as many sources as possible should be
practiced in causal inference (36).

@@

where \(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
represents the imprecision (or noise) in the measurement of \(x\). The
same model of measurement error can be applied to the outcome variable
\(y\). In this study we assume there is no measurement error in the SNP,
and that measurement error in the exposure and the outcome are
uncorrelated.

@@

\begin{itemize}
\tightlist
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
  relationship
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then the confounding model is
  accepted (\(x \leftarrow g \rightarrow y\)).
\end{itemize}

@@

Note that the same correlation test approach can be applied to a
two-sample MR (28) setting. Two-sample MR refers to the case where the
SNP-exposure association and SNP-outcome association are calculated in
different samples (e.g.~from publicly available summary statistics).
Here the Steiger test of two independent correlations can be applied
where.

@@

\subsubsection{Causal direction sensitivity
analysis}\label{causal-direction-sensitivity-analysis}

@@

To evaluate how reliable, \(R\), the inference of the causal direction
is to potential measurement error in \(x\) and \(y\) we need to predict
the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
error. We integrate over the entire range of \(\rho_{gy} - \rho_{gx}\)
values for possible measurement error values. We find the ratio of the
volume that agrees with the inferred direction of causality over the
volume that disagrees with the inferred direction of causality. A ratio
\(R=1\) indicates that the inferred causal direction is highly sensitive
to measurement error, with equal weight of the measurement error
parameter space supporting both directions of causality. In general, the
\(R\) value denotes that the inferred direction of causality is \(R\)
times more likely to be the empirical result than the opposite
direction. Full details are provided in Appendix 2.

@@

where \(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), \(\alpha_{mx}\) and
\(\beta_{mx}\) are parameters that represent non-differential
measurement error into the exposure variable \(x\), and \(\alpha_{my}\)
and \(\beta_{my}\) are parameters for non-differential measurement error
in the outcome \(y\). Similarly in the non-causal model:

\[
\begin{aligned}
g & \sim Binom(2, 0.5) \\
u & = \alpha_g + \beta_g g + \epsilon_g \\
x & = \alpha_u + \beta_u u + \epsilon_{u_x} \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y & = \alpha_u + \beta_u u + \epsilon_{u_y} \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
\end{aligned}
\]

All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
Normally distribted values of \(\epsilon_*\) were generated such that

\[
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(g, u)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
\]

@@

We performed simulations to verify that this problem does arise using
the CIT method. Figure 2 shows that when there is no measurement error
in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
reliable in identifying the correct causal direction. However, as
measurement error increases in the exposure variable, eventually the CIT
is more likely to infer a robust causal association in the wrong
direction. Also of concern here is that increasing sample size does not
solve the issue, indeed it only strengthens the apparent evidence for
the incorrect inference.

We also performed simulations to compare the performance of MR against
CIT in detecting a causal association between simulated variables under
different levels of imprecision simulated in the exposure. Figure 3
shows the true positive rates between the CIT and MR for detecting a
causal association. We observe that the CIT has lower power in all
cases, with performance declining as measurement imprecision increases
in the exposure. When MR assumptions are satisfied, notably that it is
known on which of \(x\) and \(y\) the SNP \(g\) has a primary influence,
the performance of MR in detecting an association is unrelated to
measurement error in the exposure. Measurement error in the outcome does
reduce power, but does not induce a substantive difference in
performance between CIT and MR.

@@

If we do not know whether the SNP \(g\) has a primary influence on \(x\)
or \(y\) then CIT can attempt to infer the causal direction. Here we
introduce the MR Steiger approach to similarly orient the direction of
causality but in an MR analysis when the underlying biology of the SNP
is not fully understood.

For a particular association, it is of interest to identify the range of
possible measurement error values agree and disagree with the
empirically inferred causal direction (Figure 4a, Appendix 2). This
metric can be used to evaluate the reliability of MR Steiger.

We show that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2) determines
the range of parameters around which the Steiger test is liable to
provide the wrong direction of causality (\emph{i.e.} if \(d>0\) then
the Steiger test is likely to be correct about the causal direction).
Figure 4b shows that when there is no measurement error in \(x\), the
Steiger test is unlikely to infer the wrong direction of causality even
if there is measurement error in \(y\). It also shows that in most cases
where \(x\) is measured with error, especially when the causal effect
between \(x\) and \(y\) is not very large, the sensitivity of the
Steiger test to measurement error is relatively low.

@@

We used simulations to explore the performance of the MR Steiger
approach in comparison to CIT in terms of the rate at which evidence of
a causal relationship is obtained for the correct direction of
causality, and the rate at which evidence of a causal relationship is
obtained where the reported direction of causality is incorrect.
Simulations were performed for two models, one for a ``causal model''
where there was a causal relationship between \(x\) and \(y\); and one
for a ``non-causal model'' where \(x\) and \(y\) were not causally
related, but had a confounded association induced by the SNP \(g\)
influencing a confounder variable \(u\).

Figure 5a shows that, for the ``causal model'', the MR analysis is
indeed liable to infer the wrong direction of causality when \(d < 0\),
and that this erroneous result is more likely to occur with increasing
sample size. However, the CIT is in general more fallable to reporting a
robust causal association for the wrong direction of causality. When
\(d > 0\) we find that in most cases the MR method has greater power to
obtain evidence for causality than CIT, and always obtains the correct
direction of causality. The CIT, unlike the Steiger test for MR, is able
to distinguish the ``non-causal model'' from the ``causal model''
(Methods, Figure 5b), but it is evident that measurement error will
often lead the CIT to identify the causal model as true, when in fact
the underlying model is this non-causal model.

@@

We used the Steiger test to infer the direction of causality between DNA
methylation and gene expression levels between 458 putative
associations. We found that the causal direction commonly goes in both
directions (Figure 6a), but assuming no or equal measurement error, DNA
methylation levels were the predominant causal factor
(\(p = 1.3\times 10^{-5}\)). The median reliability (R) of the 458 tests
was 3.92 (5\%-95\% quantiles 1.08 - 37.11). We then went on to predict
the causal directions of the associations for varying levels of
systematic measurement error for the different platforms. Figure 6a
shows that the conclusions about the direction of causality between DNA
methylation and gene expression are very sensitive to measurement error.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis are relatively well explored (44). Our analysis extends this to
related methods such as CIT that are widely used in predominantly 'omic
data. These methods are indeed susceptible to the same problem as
standard mediation based analysis, and specifically we show that as
measurement error in the (true) exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

@@

Figure 3: Outcomes were simulated to be causally influenced by exposures
with varying degrees of measurement imprecision applied to the exposure
variable (x axis). True positive rates (y axis) for MR and CIT were
compared for varying levels of measurement imprecision in the outcome
variable (rows of boxes) and sample sizes (columns of boxes).

@@

\subsubsection{Unmeasured confounding in the Steiger
test}\label{unmeasured-confounding-in-the-steiger-test}

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Orienting the causal relationship between imprecisely measured traits using GWAS summary data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{18 July 2017}

@@

Inference of the causal structure that induces correlations between two
traits can be achieved by combining genetic associations with a
mediation-based approach, as is done in the causal inference test (CIT).
However, we show that measurement error in the phenotypes can lead to
the CIT inferring the wrong causal direction, and that increasing sample
sizes has the adverse effect of increasing confidence in the wrong
answer. This problem is likely to be general to other mediation-based
approaches. Here we introduce an extension to Mendelian randomisation, a
method that uses genetic associations in an instrumentation framework,
that enables inference of the causal direction between traits, with some
advantages. First, it can be performed using only summary level data
from genome-wide association studies; second, it is less susceptible to
bias in the presence of measurement error or unmeasured confounding. We
apply the method to infer the causal direction between DNA methylation
and gene expression levels. Our results demonstrate that, in general,
DNA methylation is more likely to be the causal factor, but this result
is highly susceptible to bias induced by systematic differences in
measurement error between the platforms, and by horizontal pleiotropy.
We emphasise that, where possible, implementing MR and appropriate
sensitivity analyses alongside other approaches such as CIT is important
to triangulate reliable conclusions about causality.

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding (1--3). In response to this, the use of genetic
associations to instrument traits has emerged as a technique for
improving the reliability of causal inference in observational data, and
with the coincident rise in genome-wide association studies it is now a
prominent tool that is applied in several different guises (3--6).
However, shifting from observational associations to instrumentation
does require more (often untestable) assumptions, and potential pitfalls
remain. One that is often neglected is the influence of non-differential
measurement error on the reliability of causal inference.

@@

Measurement error will of course arise in other types of data too. For
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect (12), leading to the problem
that phenotypes may be imprecisely defined. A similar problem of
biological misspecification is unavoidable in disease diagnosis, and
measuring behaviour such as smoking or diet is notoriously difficult to
do accurately. Measurement error can also be introduced after the data
have been collected, for example the transformation of non-normal data
for the purpose of statistical analysis will lead to a new variable that
will typically incur both changes in scale and imprecision (noise)
compared to the original variable. The sources of measurement error are
not limited to this list (8), and its impact has been explored in the
epidemiological literature extensively (13,14). Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

@@

MR can be applied to the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP is used as a surrogate for the exposure. Assuming the SNP
associates with the outcome only through the exposure, the causal effect
of the exposure on the outcome can be estimated by scaling the
association between the SNP and the outcome by the association between
the SNP and the exposure. Though difficult to test empirically, this
assumption can be relaxed in various ways when multiple instruments are
available for a putative exposure (24,25) and a number of sensitivity
tests are now available to improve reliability (26). Additionally, if
valid genetic instruments are known for both traits of interest then MR
can be performed in both directions (bi-directional MR), testing the
influence of one trait on the other and vice versa, to infer the causal
direction between the two phenotypes.

@@

MR has some important advantages over the mediation-based approaches.
First, the mediation-based approaches require that the exposure, outcome
and instrumental variables are all measured in the same data, whereas
recent extensions to MR circumvent this requirement, allowing causal
inference to be drawn when exposure variables and outcome variables are
measured in different samples (28). This has the crucial advantage of
improving statistical power by allowing analysis in much larger sample
sizes, and dramatically expands the breadth of possible phenotypic
relationships that can be evaluated (26). Second, the mediation-based
approach of adjusting the outcome for the exposure to nullify the
association between the SNP and the outcome is affected by unmeasured
confounding of the exposure and outcome. This is because adjusting the
outcome by the exposure induces a collider effect between the SNP and
outcome (29), and in order to fully abrogate this association one must
also adjust for all (hidden or otherwise) confounders. MR does not
suffer from this problem because it does not test for association
through adjustment. Third, when MR assumptions are satisfied the method
is robust to there being measurement error in the exposure variable
(30). Indeed instrumental variable (IV) analysis was in part initially
introduced as a correction for measurement error in the exposure (31),
whereas it has been noted that both classic mediation-based analyses
(13,14,32,33) and mediation-based methods that use instrumental
variables (34,35) are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. Though we only examine the CIT method in
detail, we show that attempting to adjust for mediating variables to
make causal inference is susceptible to problems, which can be
generalised to other mediation-based methods. We then present an
extension to MR that allows researchers to ascertain the causal
direction of an association even when the biology of the instruments are
not fully understood, and also a metric to evaluate the sensitivity of
the result of this extension to measurement error. Finally, to
demonstrate the potential impact of measurement error we apply this
method to infer the direction of causation between DNA methylation
levels and gene expression levels. Our analyses highlight that because
these different causal inference techniques have varying strengths and
weaknesses, triangulation of evidence from as many sources as possible
should be practiced in causal inference (36).

@@

where \(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
represents the imprecision (or noise) in the measurement of \(x\).
Measurement imprecision can represent imprecise measurement due to
limits on sensitivity of measuring equipment, or due to phenotypes being
imprecisely defined. The same model of measurement error can be applied
to the outcome variable \(y\).

In this study we assume there is no measurement error in the SNP. Common
genetic variants are typically less susceptible to measurement error due
to strict quality control procedures prior to genome wide association
studies. Any non-differential measurement error that might be present
(either because the SNP is poorly typed or because the SNP is not in
complete linkage disequilibrium with the causal variant) will reduce
power in MR but will not incur bias (3,13,30). We also assume that
measurement error in the exposure and the outcome are uncorrelated.

@@

\begin{itemize}
\tightlist
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
  relationship
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then there is likely confounding
  and no call is made (\(x \leftarrow u \rightarrow y\)).
\end{itemize}

@@

Note that the same correlation test approach can be applied to a
two-sample MR setting. Two-sample MR refers to the case where the
SNP-exposure association and SNP-outcome association are calculated in
different samples (e.g.~from publicly available summary statistics
(26,28)). Here the Steiger test of two independent correlations can be
applied where.

@@

The Steiger test assumes that there is a causal relationship between the
two variables, and that the SNP is a valid instrument for one of them.
However it is liable to give incorrect causal directions under some
other circumstances also. First, some levels of horizontal pleiotropy,
where the SNP influences the outcome through some pathway other than the
exposure, could induce problems because this is a means by which the
instrument is invalid. Second, some differential values of measurement
error between the exposure and the outcome could induce incorrect causal
directions, and this is explored in Appendix 2. Third, some levels of
unmeasured confounding between the exposure and the outcome could induce
the wrong causal direction, and this is explored in Appendix 3.

\subsubsection{Causal direction sensitivity analysis for measurement
error}\label{causal-direction-sensitivity-analysis-for-measurement-error}

@@

To evaluate how reliable, \(R\), the inference of the causal direction
is to potential measurement error in \(x\) and \(y\) we need to predict
the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
error. We offer two tools in which to do this. First, the user can
provide values of measurement error for \(x\) and \(y\) and obtain a
revised inference of the causal direction. Second, we integrate over the
entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
measurement error values, assuming that any measurement error value is
equally likely. Across all possible values of measurement error in \(x\)
and \(y\) we find the volume that agrees with the inferred direction of
causality and the volume that disagrees with the inferred direction of
causality, and take the ratio of these two values. A ratio \(R=1\)
indicates that the inferred causal direction is highly sensitive to
measurement error, because equal weight of the measurement error
parameter space supporting both directions of causality. In general, the
\(R\) value denotes that the inferred direction of causality is \(R\)
times more likely to be the empirical result than the opposite
direction. Full details are provided in Appendix 2.

@@

where non-differential measurement error is represented by a noise
(measurement imprecision) term
\(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), and measurement bias terms
\(\alpha_{m*}\) and \(\beta_{m*}\) for the exposure variable \(x\) and
the outcome variable \(y\). Note that following the first section of the
Results we no longer include the bias terms for simplicity. We have
formulated the the non-causal model as:

\[
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \alpha_{gx} + \beta_{gx} g + \epsilon_{gx} \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y & = \alpha_{gy} + \beta_{gy} g + \epsilon_{gy} \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
\end{aligned}
\]

All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
Normally distributed values of \(\epsilon_*\) were generated such that

\[
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
\]

@@

We performed simulations to verify that this problem does arise using
the CIT method. Figure 2 shows that when there is no measurement error
in the exposure or outcome variables (\(\rho_{x, x_o}=\rho_{y, y_o}1\))
the CIT is reliable in identifying the correct causal direction.
However, as measurement error increases in the exposure variable,
eventually the CIT is more likely to infer a robust causal association
in the wrong direction. Also of concern here is that increasing sample
size does not solve the issue, indeed it only strengthens the apparent
evidence for the incorrect inference.

@@

If we do not know whether the SNP \(g\) has a primary influence on \(x\)
or \(y\) then CIT can attempt to infer the causal direction. Though
bi-directional MR can be used to orient causal directions, this requires
knowledge of a valid instrument for each trait, and we were motivated to
develop the MR Steiger method that could operate on summary data to
orient the direction of causality using the same conditions as the CIT,
where the underlying biology of a single SNP is not fully understood. We
go on to explore the scenarios in which the method is likely to return
the correct or incorrect causal directions.

We performed simulations to compare the power and type 1 error rates of
MR and CIT in detecting a causal association between simulated variables
under different levels of imprecision simulated in the exposure.
Comparing the performance of methods with different sets of assumptions
can be difficult, but a basic comparison is shown in Figure 3. We
observe that the CIT is more conservative under the null model of no
association owing to the omnibus test statistic comprising several
statistical tests. The FDR using a p-value threshold of 0.05 appears to
be close to 0, whereas for the MR Steiger method the FDR is around 0.05.
Using the same p-value thresholds to declare significance in the
non-null simulations, the general trend appears to be that the CIT power
reduces as measurement error in the exposure increases faster than that
of the MR Steiger method.

For a particular association, it is of interest to identify the range of
possible measurement error values for which the method will give results
that agree or disagree with the empirically inferred causal direction
(Figure 4a, Appendix 2). This metric can be used to evaluate the
reliability of MR Steiger test.

We show that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2) determines
the range of parameters around which the MR Steiger test is liable to
provide the wrong direction of causality (\emph{i.e.} if \(d>0\) then
the MR Steiger test is likely to be correct about the causal direction).
Figure 4b shows that when there is no measurement error in \(x\), the MR
Steiger test is unlikely to infer the wrong direction of causality even
if there is measurement error in \(y\). It also shows that in most cases
where \(x\) is measured with error, especially when the causal effect
between \(x\) and \(y\) is not very large, the sensitivity of the MR
Steiger test to measurement error is relatively low.

Unmeasured confounding between the exposure and outcome can also give
rise to problems with the MR Steiger approach (Appendix 3). Though the
relationship between unmeasured confounding and causal orientation is
complex (Supplementary Figure 1: Graph representing the unmeasured
confounding parameters that will lead to the MR Steiger test returning
the wrong causal direction. Columns of boxes represent different signed
values of the observational variance explained between \(x\) and \(y\)
(\(R^2_{xy}\)).), we note that when the magnitude of the observational
variance explained between the exposure and the outcome is below 0.2 the
MR Steiger method is unlikely to return the incorrect causal direction.

@@

We used simulations to explore the performance of the MR Steiger
approach in comparison to CIT for different levels of measurement error.
The performance was compared in terms of the rate at which evidence of a
causal relationship is obtained for the correct direction of causality,
and the rate at which evidence of a causal relationship is obtained
where the reported direction of causality is incorrect. Simulations were
performed for two models, one for a ``causal model'' where there was a
causal relationship between \(x\) and \(y\); and one for a ``non-causal
model'' where \(x\) and \(y\) were not causally related, but had a
confounded association induced by the SNP \(g\) influencing a confounder
variable \(u\).

Figure 5a shows that, for the ``causal model'', the MR analysis is
indeed liable to infer the wrong direction of causality when \(d < 0\),
and that this erroneous result is more likely to occur with increasing
sample size. However, the CIT is in general more fallable to reporting a
robust causal association for the wrong direction of causality. When
\(d > 0\) we find that in most cases the MR Steiger method has greater
power to obtain evidence for causality than CIT, and always obtains the
correct direction of causality. The CIT, unlike the MR Steiger test, is
able to distinguish the ``non-causal model'' from the ``causal model''
(Methods, Figure 5b), but it is evident that measurement error will
often lead the CIT to identify the causal model as true, when in fact
the underlying model is this non-causal model.

@@

We used the MR Steiger test to infer the direction of causality between
DNA methylation and gene expression levels between 458 putative
associations. We found that the causal direction commonly goes in both
directions (Figure 6a), but assuming no or equal measurement error, DNA
methylation levels were the predominant causal factor
(\(p = 1.3\times 10^{-5}\)). The median reliability (R) of the 458 tests
was 3.92 (5\%-95\% quantiles 1.08 - 37.11). We then went on to predict
the causal directions of the associations for varying levels of
systematic measurement error for the different platforms. Figure 6a
shows that the conclusions about the direction of causality between DNA
methylation and gene expression are very sensitive to measurement error.
We made a strong assumption that either methylation influenced gene
expression or vice versa, but it is certainly possible that the SNP is
solely or additionally influencing some other trait that confounds the
association between gene expression and DNA methylation.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis are relatively well explored (44). Our analysis extends this to
related methods such as CIT that are used in predominantly 'omic data.
These methods are indeed susceptible to the same problem as standard
mediation based analysis, and specifically we show that as measurement
error in the (true) exposure variable increases, CIT is likely to have
reduced statistical power, and liable to infer the wrong direction of
causality. We also demonstrate that, though unintuitive, increasing
sample size does not resolve the issue, rather it leads to more extreme
p-values for the model that predicts the wrong direction of causality.

@@

Figure 3: Outcomes were simulated to be unrelated to the exposure
(bottom plot) or causally influenced by the exposure (top plot) with
varying degrees of measurement imprecision applied to the exposure
variable (x axis). True positive rates (y axis) for MR and CIT were
compared for varying sample sizes (columns of boxes).

@@

\subsubsection{Unmeasured confounding in the MR Steiger
test}\label{unmeasured-confounding-in-the-mr-steiger-test}

Similarly, we can investigate the extent to which unmeasured confounding
will lead to the wrong causal direction between \(x\) and \(y\) using
the MR Steiger test, evaluating the liability for the inequality
\(cor(g,x)^2 > cor(g,y)^2\) being incorrect. After some algebra
